---
title: 'EES 3310/5310 Lab #3'
subtitle: "Exercises with the MODTRAN Model"
author: "Lauryn Gayles"
date: 'Lab: Wed. Jan. 24. Due: Wed. Jan. 31'
fontsize: "12pt"
output:
  github_document:
    toc: true
  pdf_document: 
    toc: yes
    includes:
      in_header: ees3310.sty
  word_document:
    toc: true
  html_document: 
    toc: true
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

if (knitr::is_latex_output()) {
  message("Setting PDF plot hook.")
  new_plot_hook <- function(x, options)
    paste("\n", knitr::hook_plot_tex(x, options), "\n")
  
  knitr::knit_hooks$set(plot = new_plot_hook)
} else {
  message("Not using PDF output.")
}


options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(tidyverse, knitr)

theme_set(theme_bw(base_size = 15))

source(file.path(script_dir, 'utils.R'), chdir = T)
source(file.path(script_dir, 'format_md.R'), chdir = T)
source(file.path(script_dir, 'modtran.R'), chdir = T)

```

Fill in R code for the exercises 
(I have put the comment `# TODO` in all of the code chunks where you need to 
do this)
and then fill in the answers where I have marked **Answer:**. 
Be sure to write explanations of your answer and don't just put numbers with 
no text.

## Exercise 4.1: Methane

Methane has a current concentration of 1.7 ppm in the atmosphere and
is doubling at a faster rate than CO~2~.

a) **Would an additional 10 ppm of methane in the atmosphere have a larger or 
   smaller impact on the outgoing IR flux than an additional 10 ppm of CO~2~ 
   at current concentrations?**

   **Hint:**
   See the suggestion in the `lab-03-instructions` document.

```{r ex_4_1_a, include=TRUE, message=FALSE}

modtran_400_17.txt = run_modtran("my_modtran_file.txt", co2_ppm = 400, ch4_ppm = 1.7) #run with different values
modtran_410_17.txt = run_modtran("my_modtran_file.txt", co2_ppm = 410, ch4_ppm = 1.7) #run with different values
modtran_400_117.txt = run_modtran("my_modtran_file.txt", co2_ppm = 400, ch4_ppm = 11.7) #run with different values



modtran_400_17 <- modtran_400_17.txt$i_out #isolate the IR flux 
print(modtran_400_17) #print 

modtran_410_17 <- modtran_410_17.txt$i_out #isolate the IR flux 
print(modtran_410_17) #print 

modtran_400_117 <- modtran_400_117.txt$i_out #isolate the IR flux 
print(modtran_400_117) #print 


```

**Answer:** _Put your answer here._ Be sure to explain your reasoning and 
show data, plots, etc. to explain how you came up with your answer.

The additional 10 ppm of methane has a larger impact than an additional 10 ppm of CO2. We can see the decrease in the IR flux was much greater with the additional methane than the additional CO2. 


b) **Where in the spectrum does methane absorb? What concentration does it take 
   to begin to saturate the absorption in this band? Explain what you are looking 
   at to judge when the gas is saturated.**

   **Hints:**  
   See the hints in the `lab-03-instructions` document.

```{r ex_4_1_b, include=TRUE, message=FALSE}


ch4_list = c(0,1, 2, 4, 8, 16, 32, 64, 128) #creating a list to go through the different numbers

for (ch4 in ch4_list) {
mod_data = run_modtran(co2_ppm = 0, ch4_ppm = ch4, trop_o3_ppb = 0, strat_o3_scale = 0, h2o_scale = 0, freon_scale = 0)
p = plot_modtran(mod_data, descr = str_c(ch4, 'ppm ch4')) #setting the the variables to 0 
plot(p) # you could also say print(p) here.
}





```

**Answer:** 


1300 cm-1 is the wavelength where methane absorbs. The ppm for methane seems to be around 16 ppm. 
  
c) **Would a doubling of methane have as great an impact on the heat balance as a 
   doubling of CO~2~?**

   **Hint:**
   See the suggestion in the `lab-03-instructions` document.

```{r ex_4_1_c, include=TRUE, message=FALSE}

m = run_modtran("my_modtran_file.txt")
m1 = run_modtran("my_modtran_file.txt", ch4_ppm = 3.4)
m2 = run_modtran("my_modtran_file.txt", co2_ppm = 800, ch4_ppm = 1.7)



modtran <- m$i_out #isolate the IR flux 
print(modtran_400_17) #print 

modtran_1 <- m1$i_out #isolate the IR flux 
print(modtran_410_17) #print 

modtran_2 <- m2$i_out #isolate the IR flux 
print(modtran_400_117) #print 

plot_modtran(m, descr = "Baseline spectrum")
plot_modtran(m1, descr = "Doubled CO2")
plot_modtran(m2, descr = "Doubled CH4")
 



```

**Answer:** 

Doubling CO2, wavelength decreases by -3.3 W/m2. Doubling CH4, wavelength decreases by -0.88 W/m2. The increase in CO2 has a larger effect because there is more CO2 present in the atmosphere.

d) **What is the "equivalent CO~2~" of doubling atmospheric methane? That is to 
   say, how many ppm of CO~2~ would lead to the same change in outgoing IR 
   radiation energy flux as doubling methane? What is the ratio of ppm CO~2~ 
   change to ppm methane change?**

```{r ex_4_1_d, include=TRUE, message=FALSE}

matching_methane = 13 # the ppm after trial and error 
modtran_match_ch4 = run_modtran(file.path(data_dir, "ex_4_1_ch4_match.txt"),
                                co2_ppm = 400, ch4_ppm = matching_methane)

print(modtran_match_ch4) #priting data to see teh results 

```

**Answer:** _

The answer is 13 ppm which is achieved thorugh trial and error. the iout is 295.3411. 


## Exercise 4.3: Water vapor

Our theory of climate presumes that an increase in the temperature at ground
level will lead to an increase in the outgoing IR energy flux at the top of the
atmosphere.


a) **How much extra outgoing IR would you get by raising the temperature of the 
   ground by 5&deg;C? What effect does the ground temperature have on the 
   shape of the outgoing IR spectrum and why?**
   
   **Hint:**
   See the hint in the `lab-03-instructions` document.

```{r ex_4_3_a, include=TRUE, message=FALSE}

modtran_baseline = run_modtran() #run the baseline modtran 
modtran_plus_5 = run_modtran(delta_t = 5) #change the temperature variable 

p_baseline = plot_modtran(modtran_baseline, descr = "Baseline spectrum") #plot
p_5_deg = plot_modtran(modtran_plus_5, #plot 
                        descr = "Ground temperature raised 5K")
plot(p_baseline) #print the plot 

```

**Answer:** 

We can see in the grpah that an increase in the ground temperature causes a raise in the spectrum as a whole. 

b) **More water can evaporate into warm air than into cool air. Change the
   model settings to hold the water vapor at constant relative humidity 
   rather than constant vapor pressure (the default), calculate the change
   in outgoing IR energy flux for a 5&deg;C temperature increase.
   Is it higher or lower? Does water vapor make the Earth more sensitive to
   CO~2~ increases or less sensitive?**
   
   **Note:** By default, the MODTRAM model holds water vapor pressure constant,
   but you can set it to hold relative humidity constant instead with the
   option `h2o_fixed = "relative humidity"`, like this:
   `run_modtran(file_name, delta_t = 5, h2o_fixed = "relative humidity")`.

```{r ex_4_3_b, include=TRUE, message=FALSE}

modtran_vp = run_modtran(delta_t = 5, h2o_fixed = "vapor pressure") #run modtran with temp change being 5 
modtran_rh = run_modtran(delta_t = 5, h2o_fixed = "relative humidity") #run modtran with temp change being 5 

i_base = modtran_baseline$i_out #getting the i out data 
i_water_vapor = modtran_vp$i_out
i_humidity = modtran_rh$i_out

p_water_vapor = plot_modtran(modtran_vp, descr = "Constant vapor pressure", 
                    i_out_ref = i_base)
p_humidity = plot_modtran(modtran_rh, descr = "Constant relative humidity", 
                    i_out_ref = i_base)
plot(p_water_vapor)
plot(p_humidity) #plotting the data 


```

**Answer:** 

The increase in temp has an increase effect on iout when the fixed variable is water vapor. The environment is more sensitive to co2 changing when the humidity remain the same.  


c) **Now see this effect in another way.**

   * **Starting from the default base case, record the total outgoing 
     IR flux.**

   * **Now double CO~2~. The temperature in the model stays the
     same (that's how the model is written), but the outgoing IR flux
     goes down.**

   * **Using constant water vapor pressure, adjust the temperature offset
     until you get the original IR flux back again. Record the change in
     temperature.**
    
   * **Now repeat the exercise, but holding the relative humidity fixed
     instead of the water vapor pressure.**
    
   * **The ratio of the warming when you hold relative humidity fixed
     to the warming when you hold water vapor pressure fixed is the 
     feedback factor for water vapor. What is it?**

```{r ex_4_3_c, include=TRUE, message=FALSE}


i_base = modtran_baseline$i_out

modtran_800co2 = run_modtran(co2_ppm = 800) #run modtran with co2 at 800 ppm 
i_vapor_double = modtran_800co2$i_out #get the iout value 

dt_vp = 0.76
modtran_vp_dt = run_modtran(file.path(data_dir, "ex_4_3_vp_dt.txt"),
                            co2_ppm = 800, delta_t = dt_vp) #run modtran with co2 at 800 ppm
i_vp_dt = modtran_vp_dt$i_out #get the iout value

modtran_2x_rh = run_modtran(file.path(data_dir, "ex_4_3_2x_co2_rh_.txt"),#run modtran with co2 at 800 ppm
                            co2_ppm = 800, h2o_fixed = "relative humidity")
i_2x_rh = modtran_2x_rh$i_out

dt_rh = 1.21
modtran_rh_dt = run_modtran(file.path(data_dir, "ex_4_3_rh_dt.txt"),
                            co2_ppm = 800, delta_t = dt_rh, 
                            h2o_fixed = "relative humidity")
i_rh_dt = modtran_rh_dt$i_out

feedback = dt_rh / dt_vp

print(feedback) # print to see all values 

```

**Answer:** _Put your answer here._ 

The answer is 1.592 for the feedback factor. This shows the the ratio of the temp change with relative humidity fixed to that of the temp change with water vapor fixed. 
